Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-04 Paper/manuscript/manuscript_05.tex
section 3 ready to get feedback

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of selecting a subset $m$ of training points on which to base computation. They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. New developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.}

@@

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\mathbf{x}-\mathbf{x}'$, such that it can be written $k(\mathbf{x},\mathbf{x}') = k(\mathbf{x}-\mathbf{x}')$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(\mathbf{r})$, which means that the covariacne is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which are
%
\begin{eqnarray}
k_{\nu=\infty}(\mathbf{r})&=&\sigma^2 \text{exp}(-\frac{1}{2} \mathbf{r}/\ell^2), \nonumber \\
k_{\nu=\frac{1}{2}}(\mathbf{r})&=&\sigma^2 exp(-\mathbf{r}/\ell), \nonumber \\
k_{\nu=\frac{3}{2}}(\mathbf{r})&=&\sigma^2(1+\sqrt{3}\mathbf{r}/\ell \text{exp}(-\sqrt{3}\mathbf{r}/\ell), \nonumber \\
k_{\nu=\frac{5}{2}}(\mathbf{r})&=&\sigma^2(1+\sqrt{5}\mathbf{r}/\ell+\frac{5}{3}\mathbf{r}^2/\ell^2) \text{exp}(-\sqrt{5}\mathbf{r}/\ell), \nonumber 
\end{eqnarray}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the lengt-scale and magnitud, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to the use of multidimensional length-scale $\boldsymbol{\ell}$, which basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{eqnarray}
S_{\nu=\infty}(\mathbf{w})&=& \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \text{exp}(-0.5 \ell^2 \mathbf{w}^2), \nonumber \\
S_{\nu=\frac{1}{2}}(\mathbf{w})&=& 2\sigma^2 \frac{1}{\ell}(\frac{1}{\ell^2} + \omega^2)^{-1}, \nonumber \\
S_{\nu=\frac{3}{2}}(\mathbf{w})&=& 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \omega^2)^{-2}, \nonumber \\
S_{\nu=\frac{5}{2}}(\mathbf{w})&=& \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}(\frac{5}{\ell^2} + \omega^2)^{-3}, \nonumber 
\end{eqnarray}

\noindent where variable $\mathbf{w}$ is in the frequency domain.

@@

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function. For example, the $D$-dimensional spectral density functions associated with the Mat\'ern covariance functions considered in the previous section are
%
\begin{eqnarray}
S^{\ast}_{\nu=\infty}(\mathbf{w})&=& \sigma^2 \sqrt{2\pi}^D \prod_{d=1}^D l_d \cdot \text{exp}\left(-0.5 \sum_{d=1}^D l_d^2 \mathbf{w}_d^2\right), \nonumber \\
%
S^{\ast}_{\nu}(\mathbf{w})&=& \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\mathbf{w}^2 \right)^{(\nu+D/2)}, \nonumber
\end{eqnarray}

\noindent where the $D$-vector $\mathbf{w}$ is in the frequency domain (Paul: Could it make sense to move these specific spectral density functions to Section 2.1?). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf}
f(\mathbf{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\lambda^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\mathbf{x}) \beta_j, 
\end{equation}

@@

Implementation and performance of the approximate GP model depends on the number of basis functions and box size settings, which are also related each other. The time of computation in the approximation mainly depends on the number of basis functions. At the same time, the appropriate values for the number of basis functions and box size depend on the non-linear functional effects of the function to be learned. In GP functions these non-linear functional effects are characterized by the lengthscale of the covariance function.

In this section we analyze the rule and relationships of the number of basis functions and the box size on approximation and their effects on the learned functions.

We also analyze these two factors in relation to the lengthscale of the functions to be learned. We provide a useful graph that retains how these factors, number of basis fucntions, box size and lengthscale, relate each other in relation to the performance of the model. This graph basically provides the recommended values for these factors in order to get a close approximation. This graph also allows us to diagnose an actual fit.

In this section we also make a comparison between regular GP model and approximate GP model fitted for different dataset with different "wigglyness". The comparison is made through comparing their estimated lengthscales.

@@

The approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, e.g. in a 1-$D$ input space $x\in \Omega=[-L,L]$:
%
\begin{equation}\label{diffcov}
k(\tau) \approx \sum_{j} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

\noindent where $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are those in equation (\ref{bf&lambda}) for the unidimensional case and those in equation (\ref{bf&lambda_multi}) for the multidimensional case.

The approximation to the covariance function will become exact in the limit when the number of basis functions approach infinity, $j=1,\cdots,\infty$. The number of basis functions can be truncated at some finite positive value $J$ such that the difference between the densities of the exact and approximate covariance functions be less than a threshold $\epsilon$,
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau)d(\tau) - \int \sum_{j}^J S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) d(\tau) < \epsilon.
\end{eqnarray}

The finite number $J$ of basis functions in the approximation needed to satisfy eq. (\ref{diff_covs}) will depend on the non-linear effects of the function to be learned, that is, on its lengthscale $\ell$. The approximation will also depends on the box size $L$ (equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})) which will affect the performance specially near the boundaries. The box size will also have an effect on the number of basis functions needed in the approximation. Note that we refer to the box size as the value $L$ of the Dirichelt boundary condition in eq. \ref{eigenproblem}. The box size $L$ is set as an extension of the desired predicting input domain $\Psi$. Considering a zero-mean input domain $\Psi=[-S,S] \in \mathbb{R}$, the box size is defined as,
%
\begin{equation}
L=c \cdot S
\end{equation}

\noindent where $S$ represents the half-range of the input space, and $c$ ($c \ge 1$) is the proportional extension factor. From now on we are going to refer to this proportional factor $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the box size $L$ normalized by the half-range of the input space $S$.\\

For a naive illustration of the effects of these two factors, number of basis functions $J$ and box size $L$, on the performance of the approximation, a set of noisy observations are simulated from a GP model with lengthscale $\ell=0.3$ and magnitud $\alpha=1$ for input vales within the input domain with half-range $S=1$. Approximate GP models with different number of basis functions and box sizes $L$ ($L=c\cdot S$, with $c\ge 1$) are fitted to this data. The lengthscale and magnitud parameters are considered fix to the true values of the generative GP model. Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} ilustrate the individual effects of the number of basis functions and the box size, respectively, on the approximation of the covariance function. On the one hand, for a fixed $L$, let's suppose it is set to a right choice, Figure \ref{fig1_Post_J} shows the effect of changing the number of basis functions in approaching the posterior mean and the covariance function. It can be seen how the number of basis functions affects the non-linear effects of the posteriors. Fewer basis functions implies smoother covariance functions and consequently smoother posteriors. The more "wigglyness" of the function, the more basis functions are needed. On the other hand, and similarly, for a fixed number of basis functions $J$, let's suppose it is set to a right choice, Figure \ref{fig2_Post_L} shows the effect of changing the box size $L$ in approaching the posterior mean and the covariance function. It can be seen how the box size mainly affects the approximation near the boundaries of the function and affects the covariances at long distances. \\

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.43, trim = 0mm 0mm 0mm 10mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.43, trim = 0mm 0mm 0mm 10mm, clip]{fig1_Cov_J.pdf}}
\caption{(left) Posterior distributions for different number of basis functions $J$. (right) Covariance function for different number of basis functions $J$.}
  \label{fig1_Post_J}
\end{figure}

@@

In the previous Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} we focused on illustrating the individual effects of the number of basis functions and the box size. Furthermore, for a clearer illustration of these effects, the lengthscale and magnitud parameters in the approximate GP model were fixed to their true values. Now we are going to focus on analyzing the interaction effects between these two factors, number of basis functions and the boundary factor, on the performance of the approximation. The lengthscale and magnitud parameters will not be fixed rather they will be estimated in both regular and approximate GP models.  
%The right-side plots in Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} are computed using the lengthscale and magnitud values of the true GP prior. Now, we focus on the estimated lengthscale and magnitud parameters after fitting the model, either of the regular GP model fit and approximate GP model fit. 
Figure \ref{fig4_Post_&_Cov_part1} shows the posteriors means and the covariance functions obtained after fitting the data, for different number of basis functions and boundary factors. Figure \ref{fig5_MSE_vs_J} shows the root mean square error (RMSE) of the approximate GP model over the regular GP model as a function of the number of basis functions and the boundary factors. Figure \ref{fig6_lscale_vs_J} shows the lengthscale and magnitud parameters for the regular GP model and for the approximate models as a function of the number of basis functions and boundary factors. Looking at the RMSEs in Figure \ref{fig5_MSE_vs_J}, it can be drawn that the optimal choice in terms of precision and computations would be 15 basis functions and a boundary factor between 1.5 and 2.5. The choice of 10 basis functions and a boundary factor of 1.5 could be an accurate enough choice. This same conclusion can be also be intuitively seen in the posterior means and covariance functions plots in Figure \ref{fig4_Post_&_Cov_part1}. \\

From Figure \ref{fig5_MSE_vs_J}, a behaviour in terms of performance of the approximate GP model as a function of the number of basis functions and the boundary factor, can be deduced: \\
$\bullet$ as the boundary factor increases, more basis functions are needed,
\\
$\bullet$ as fewer basis functions are used, the boundary factor must decrease. \\

Similar conclusions can be stated from results of Figure \ref{fig6_lscale_vs_J}, This Figure shows the estimated lengthscale and magnitud parameter as function of the number of basis functions and the boundary factor. In addition to the above conclusions, it can be easily seen that exists a minimum value for the boundary factor under which a close approximation will never be achieved. \\

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.6]{fig4_Post_&_Cov_part1.pdf}}
\subfigure{\includegraphics[scale=0.6]{fig4_Post_&_Cov_part2.pdf}}
\caption{Posterior distributions and covariance functions of the proposed approximate GP models and the exact GP model in function of the number of basis functions $J$ and the boundary factor $c$.} 
  \label{fig4_Post_&_Cov_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4]{fig5_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig5_MSE_vs_c.pdf}}
\caption{Standarized root mean square error (SRMSE) of the proposed approximate GP models and the exact GP model. (left) SRMSE against the number of basis functions $J$ for different values of the boundary factor $c$. (right) SRMSE against the boundary factor $c$ for different values of the number of basis functions $J$. }
  \label{fig5_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4]{fig6_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig6_magnitud_vs_J.pdf}}
\caption{(left) Estimated lengthscales against the number of basis functions $J$ for different values of the boundary factor $c$. (right) Estimated GP magnitud the number of basis functions $J$ for different values of the boundary factor $c$. }
  \label{fig6_lscale_vs_J}
\end{figure}

Additionally, there exit a relation of the number of basis functions $J$ and the boundary factor $c$ with the lengthscale $\ell$ of the function. Figure \ref{fig3_lscale_vs_J_vs_c_part1} collects how these three factors relate each other in relation to the performance of the approximation evaluated using eq. (\ref{diff_covs}). On the X-axis of this plot is placed the lengthscale of the process normalized by the half-range of the data $\ell/S$. The countour lines gather the boundary factor $c$. And the Y-axis represents the number of basis functions. Thus, this figure give us, for a certain GP function with lengthscale $\ell$ and given a fixed boundary factor $c$, the minimum number of basis functions needed to achieve a close approximation in terms of satisfying eq. (\ref{diff_covs}). Alternatively, this figure could also be read as the minimum boundary factor $c$ that we should use given a number of basis functions, for certain GP function with certain lengthscale. And also this figure could also be read as the minimum functional lengthscales that can be closely approximated given a number of basis functions and a boundary factor. This figure basically collects the behaviour of the approximate GP model as a function of the lengthscale, number of basis functions and boundary factor:

@@

$\bullet$ This plot serves as a diagnosis tool in the sense that if a estimated lengthscale is lower than that minimun that the figure bring us, given certain number of basis functions and certain boundary factor, means that the number of basis functions should be increased or the boundary factor decreased. On the other hand, if the lengthscale is bigger than that generated by the figure means that the approximation should be close enough.\\

Figure \ref{fig3_lscale_vs_J_vs_c_part1} is useful for the user to know the suitable values for the number of basis functions $J$ and the boundary factor $c$ to approximate certain process characterized by its lengthscale. This Figure will allow us to optimize computations which depends on the number of basis functions. Starting from some guess of the lengthscale of the process and knowing the range of the input data which we are interested in, the user can do a few iterative adjustments to obtain an optimal fit. This figure also provides a diagnosis tool of the fit by comparing a actual estimated lengthscale to that the figure bring.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.4]{fig3_lscale_vs_J_vs_c_part1.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig3_lscale_vs_J_vs_c_part2.pdf}}
\caption{Relation between the minumum number of basis functions $J$, the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), and the boundary condition factor $c$ ($c = \frac{L}{S}$).}
  \label{fig3_lscale_vs_J_vs_c_part1}
\end{figure}

If we look back to the conclusions drawn from Figures \ref{fig5_MSE_vs_J} and \ref{fig6_lscale_vs_J} where 10 basis functions and a boundary factor of 1.5 would be enough to closely approximate a function with normalized lengthscale eaual to 0.3, we can recognize that these conclusions also matches the conclusion brought form Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

@@

Finally, we make a comparative analysis between the lengthscales estimated by a regular GP model and approximate GP model in several datasets with different "wigglyness" (or lengthscales). From this analysis we can recognize when an approximation is not close enough to the regular GP when the lengthscales are not similar. We contrast these results from these lengthscale similarities between the regular and the approximate GP model with the information brought from the Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

>>>>

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of selecting a subset $m$ of training points on which to base computation. They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. New developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

@@

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\mathbf{x}-\mathbf{x}'$, such that it can be written $k(\mathbf{x},\mathbf{x}') = k(|\mathbf{x}-\mathbf{x}'|)$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(\mathbf{r})$, which means that the covariacne is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which, in the unidimensional case (i.e. $\mathbf{x}=x\in \mathbb{R}$, so $\mathbf{r}=r\in \mathbb{R}$), are
%
\begin{eqnarray}
k_{\nu=\infty}(r)&=&\sigma^2 \text{exp}(-\frac{1}{2} r^2/\ell^2), \nonumber \\
k_{\nu=\frac{1}{2}}(r)&=&\sigma^2 exp(-r/\ell), \nonumber \\
k_{\nu=\frac{3}{2}}(r)&=&\sigma^2(1+\sqrt{3}r/\ell \text{exp}(-\sqrt{3}r/\ell), \nonumber \\
k_{\nu=\frac{5}{2}}(r)&=&\sigma^2(1+\sqrt{5}r/\ell+\frac{5}{3}r^2/\ell^2) \text{exp}(-\sqrt{5}r/\ell), \nonumber 
\end{eqnarray}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the lengt-scale and magnitud, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to the multidimensional case, with multidimensional inputs $\mathbf{x}\in \mathbb{R}^D$ and multidimensional length-scale $\boldsymbol{\ell}\in \mathbb{R}^D$ (Gabi: We should derive the multidimensional form of the previous kernels, but I have now doubts how is the multidimensional form for the Matern kernels with $\nu$=1/2, $\nu$=3/2 and $\nu$=5/2). Using a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{eqnarray}
S_{\nu=\infty}(w)&=& \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \text{exp}(-0.5 \ell^2 w^2), \\
S_{\nu=\frac{1}{2}}(w)&=& 2\sigma^2 \frac{1}{\ell}(\frac{1}{\ell^2} + \omega^2)^{-1}, \\
S_{\nu=\frac{3}{2}}(w)&=& 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \omega^2)^{-2}, \\
S_{\nu=\frac{5}{2}}(w)&=& \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}(\frac{5}{\ell^2} + \omega^2)^{-3},  
\end{eqnarray}

\noindent where variable $w$ is in the frequency domain. The $D$-dimensional spectral density functions associated with the Mat\'ern covariance functions considered in the previous section are
%
\begin{eqnarray}
S^{\ast}_{\nu=\infty}(\mathbf{w})&=& \sigma^2 \sqrt{2\pi}^D \prod_{d=1}^D l_d \cdot \text{exp}\left(-0.5 \sum_{d=1}^D l_d^2 \mathbf{w}_d^2\right), \nonumber \\
%
S^{\ast}_{\nu}(\mathbf{w})&=& \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\mathbf{w}^2 \right)^{(\nu+D/2)}, \nonumber
\end{eqnarray}

\noindent where the $D$-vector $\mathbf{w}$ is in the frequency domain.

@@

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf}
f(\mathbf{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\lambda^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\mathbf{x}) \beta_j, 
\end{equation}

@@

Implementation and performance of the HSGP model depends on the number of basis functions and boundary settings, which are, at the same time, related  to each other. The time of computation in the approximation mainly depends on the number of basis functions. Furthermore, the appropriate values for the number of basis functions and boundary depend on the non-linear functional effects of the function to be learned. In GP functions these non-linear functional effects are characterized by the lengthscale of the covariance function.

In this section we analyze the rule and relationships of the number of basis functions and boundary on the approximation and their effects on the learned functions. 
We also analyze these two factors in relation to the lengthscale of the functions to be learned. We provide a useful graph showing how these factors, number of basis functions, boundary and lengthscale, relate to each other in relation to the performance of the model. This graph basically provides the recommended values for these factors in order to get a close approximation. This graph also allows us to diagnose an actual fit by checking whether the value of the estimated lengthscale does not keep to those that the Figure brings as suitable ones.

In this section we also make a comparison between regular GP and HSGP models, fitted for different dataset with different "wigglyness". The comparison is made through comparing their estimated lengthscales.

@@

The approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, e.g. in a 1-$D$ input space $x\in \Omega=[-L,L]\subset \mathbb{R}$:
%
\begin{equation}\label{diffcov}
k(\tau) = \sum_{j}^{\infty} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

\noindent where $L$ is the boundary, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are those in equation (\ref{bf&lambda}) for the unidimensional case and those in equation (\ref{bf&lambda_multi}) for the multidimensional case.

The approximation to the covariance function will become exact in the limit when the number of basis functions approach infinity, $j=1,\cdots,\infty$. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions be less than a threshold $\epsilon$,
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau)d(\tau) - \int \sum_{j}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) d(\tau) < \epsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy eq. (\ref{diff_covs}) will depend on the non-linear effects of the function to be learned, that is, on its lengthscale $\ell$. The approximation will also depends on boundary $L$ (equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})) which will affect the performance specially near the boundaries. The boundary will also have an effect on the number of basis functions needed in the approximation. The boundary $L$ is set as an extension of the desired predicting input domain $\Psi$. Considering a zero-mean input domain $\Psi=[-S,S] \in \mathbb{R}$, the boundary is defined as
%
\begin{equation}
L=c \cdot S,
\end{equation}

\noindent where $S$ represents the half-range of the input space, and $c$ ($c \ge 1$) is the proportional extension factor. From now on we are going to refer to this proportional factor $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by the half-range $S$ of the input space.

For a naive illustration of the effects of these two factors, number of basis functions $m$ and boundary $L$, on the performance of the approximation, a set of noisy observations are drawn from a GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, and input vales within the zero-mean input domain with half-range $S=1$. HSGP models with different number of basis functions and boundaries are fitted to this data. The lengthscale and marginal variance parameters are considered fixed to the true values of the data-generative GP model. Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} ilustrate the individual effects of the number of basis functions and the boundary, respectively, on the approximation of the covariance function. On the one hand, for a fixed $L$, let's suppose it is set to a right choice, Figure \ref{fig1_Post_J} shows the effect of changing the number of basis functions in approaching the posterior mean and the covariance function. It can be seen how the number of basis functions affects the non-linear effects of the posteriors. Fewer basis functions implies smoother covariance functions and consequently smoother posteriors. The more "wigglyness" of the function, the more basis functions are needed. On the other hand, and similarly, for a fixed number of basis functions $m$, let's suppose it is set to a right choice, Figure \ref{fig2_Post_L} shows the effect of changing the boundary $L$ in approaching the posterior mean and the covariance function. It can be seen how the box size mainly affects the approximation near the boundaries of the function and affects the covariances at long distances. \\

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.43, trim = 0mm 0mm 0mm 10mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.43, trim = 0mm 0mm 0mm 10mm, clip]{fig1_Cov_J.pdf}}
\caption{(left) Posterior distributions for different number of basis functions $m$. (right) Covariance function for different number of basis functions $m$.}
  \label{fig1_Post_J}
\end{figure}

@@

In the previous Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} we focused on illustrating the individual effects of the number of basis functions and the boundary. Furthermore, for a clearer illustration of these effects, the lengthscale and marginal variance parameters in the HSGP model were fixed to their true values. Now we are going to focus on analyzing the interaction effects between these two factors, number of basis functions and the boundary factor, on the performance of the approximation. The lengthscale and marginal variance will not be fixed rather they will be estimated either in regular and HSGP models.  
%The right-side plots in Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} are computed using the lengthscale and magnitud values of the true GP prior. Now, we focus on the estimated lengthscale and magnitud parameters after fitting the model, either of the regular GP model fit and approximate GP model fit. 
Figure \ref{fig4_Post_&_Cov_part1} shows the posteriors means and the covariance functions obtained after fitting the data, for different number of basis functions and boundary factor. Figure \ref{fig5_MSE_vs_J} shows the root mean square error (RMSE) of the approximate GP model, computed against the regular GP model, as a function of the number of basis functions and the boundary factor. Figure \ref{fig6_lscale_vs_J} shows the lengthscale and marginal variance for the regular GP model and for the HSGP models as a function of the number of basis functions and boundary factor. Looking at the RMSEs in Figure \ref{fig5_MSE_vs_J}, it can be drawn that the optimal choice in terms of precision and computations would be 15 basis functions and a boundary factor between 1.5 and 2.5. The choice of 10 basis functions and a boundary factor of 1.5 could be an accurate enough choice. This same conclusion can be also be intuitively seen in the posterior mean and covariance function plots in Figure \ref{fig4_Post_&_Cov_part1}.

From Figure \ref{fig5_MSE_vs_J}, a behaviour in terms of performance of the HSGP model as a function of the number of basis functions and the boundary factor, can be deduced: \\
$\bullet$ as the boundary factor increases, more basis functions are needed,
\\
$\bullet$ as fewer basis functions are used, the boundary factor must decrease.

Similar conclusions can be stated from results of Figure \ref{fig6_lscale_vs_J}. This Figure shows the estimated lengthscale and marginal variance as function of the number of basis functions and boundary factor. In addition to the above conclusions, in Figure \ref{fig6_lscale_vs_J} it can be easily seen that exists a minimum value for the boundary factor under which a close approximation will never be achieved. \\

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.6]{fig4_Post_&_Cov_part1.pdf}}
\subfigure{\includegraphics[scale=0.6]{fig4_Post_&_Cov_part2.pdf}}
\caption{Posterior distributions and covariance functions of the regular GP model and the proposed HSGP model as a function of the number of basis functions $m$ and the boundary factor $c$.} 
  \label{fig4_Post_&_Cov_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4]{fig5_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig5_MSE_vs_c.pdf}}
\caption{Standarized root mean square error (SRMSE) of the proposed HSGP models computed against the regular GP model. (left) SRMSE vs the number of basis functions $m$ and for different values of the boundary factor $c$. (right) SRMSE vs the boundary factor $c$ and for different values of the number of basis functions $m$. }
  \label{fig5_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4]{fig6_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig6_magnitud_vs_J.pdf}}
\caption{Estimated lengthscale and marginal variance parameters of the proposed HSGP models and the regular GP model. (left) Estimated lengthscales vs the number of basis functions $m$ and for different values of the boundary factor $c$. (right) Estimated marginal variance vs the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig6_lscale_vs_J}
\end{figure}

Additionally, there exit a relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the function. Figure \ref{fig3_lscale_vs_J_vs_c_part1} collects how these three factors relate each other in relation to the performance of the approximation evaluated using eq. (\ref{diff_covs}). On the X-axis of this plot is placed the lengthscale of the process normalized by the half-range of the data $\ell/S$. The countour lines gather the boundary factor $c$. And the Y-axis represents the number of basis functions. Thus, this figure give us, for a certain GP function with lengthscale $\ell$ and given a boundary factor $c$, the minimum number of basis functions needed to achieve a close approximation in terms of satisfying eq. (\ref{diff_covs}). Alternatively, this figure could also be read as the minimum boundary factor $c$ that we should use given a number of basis functions, for certain GP function with certain lengthscale. And also this figure could also be read as the minimum functional lengthscales that can be closely approximated given a number of basis functions and a boundary factor. This figure basically collects the behaviour of the HSGP model as a function of the lengthscale, number of basis functions and boundary factor:

@@

$\bullet$ This plot serves as a diagnosis tool in the sense that if a estimated lengthscale is lower than that minimun that the figure bring us, given certain number of basis functions and certain boundary factor, means that the number of basis functions should be increased or the boundary factor decreased. On the other hand, if the lengthscale is bigger than that generated by the figure means that the approximation should be close enough.

Figure \ref{fig3_lscale_vs_J_vs_c_part1} is useful for the user to know the suitable values for the number of basis functions $m$ and the boundary factor $c$ to approximate certain process characterized by its lengthscale. This Figure will allow us to optimize computations which depends on the number of basis functions. Starting from some guess of the lengthscale of the process and knowing the range of the input data which we are interested in, the user can do a few iterative adjustments to obtain an optimal fit. This figure also provides a diagnosis tool of the fit by comparing a actual estimated lengthscale to that the figure bring.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.4]{fig3_lscale_vs_J_vs_c_part1.pdf}}
\subfigure{\includegraphics[scale=0.4]{fig3_lscale_vs_J_vs_c_part2.pdf}}
\caption{Relation between the minumum number of basis functions $m$, the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), and the boundary factor $c$ ($c = \frac{L}{S}$).}
  \label{fig3_lscale_vs_J_vs_c_part1}
\end{figure}

If we look back to the conclusions drawn from Figures \ref{fig5_MSE_vs_J} and \ref{fig6_lscale_vs_J} where 10 basis functions and a boundary factor of 1.5 would be enough to closely approximate a function with normalized lengthscale equal to 0.3, we can recognize that these conclusions also matches the conclusion brought from Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

@@

We make a comparative analysis between the lengthscales estimated by the regular GP model and the approximate GP model in several datasets with different "wigglyness" (or lengthscales). Basically, comparing the lengthscales we can recognize if an approximation is close enough to the regular GP. We contrast these results with the information brought from Figure \ref{fig3_lscale_vs_J_vs_c_part1}.
